Compositional Certified Resource Bounds

### 2014-07-01 paper/main.tex
some updates in section 5

\section{Syntax and Semantics}

@@

\paragraph{Syntax}

@@

\paragraph{Semantics}

@@

We generalize classic Hoare logic to express not only classical
boolean-valued assertions but also assertions that talk about the
future resource usage.  Instead of the usual assertions $P : \State
\to \mathit{bool}$ of Hoare logic we use assertions
$$
P : \State \to \Q \cup \{ \infty \} \; .
$$
This can be understood as a refinement of boolean assertions where
$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Q$.
We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
$\lambda \state . \, \infty$.  To use Coq's support for propositional
reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
in the implementation.  For a given $\state \in \State$, such an
assertion can be seen as a set $B \subseteq \Q$ of valid bounds.
However, we find the presentation in this article easier to read.

@@

\paragraph{Rules of the Quantitative Logic}

@@

\paragraph{Logical Variables and Shallow Embedding}

@@

In this section we present the deductive system used in our practical
implementation to derive linear bounds on C programs.

@@

To find derivations in our quantitative logic automatically, the main
insight is to consider only assertions that match a certain shape.
The general form of a potential function that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
$$

To simplify the references to the linear coefficients $q$, we introduce
an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
\{(x, y) \mid x, y \in \Var \}$.  Each index $i$ corresponds to a \emph{base
function} $f_i$ in the potential function: $0$ corresponds to the constant
function $\_ \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can rewrite the above
equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We will often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense)
for potential functions, this allows us to represent uniquely any
linear potential function $\Phi$ as a \emph{quantitative annotation}
$Q = (q_i)_{i \in I}$, that is a family of rational numbers where
only a finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

@@

Quantitative annotations alone are not enough to make our system
functional.  A minimal logical state needs to be maintained to
justify certain operations on quantitative annotations.  For
example when analyzing the code $x \gets x + y$, it is helpful
to know the sign of $y$ to determine which intervals will increase
or decrease.  The knowledge needed by our rules can often be
inferred by local reasoning within usual theories (e.g. Presburger
arithmetic or bit vectors).

Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we designed and found
in the literature.

In contrast to the quantitative annotations, logical contexts in
the presented system are left abstract.  This allows for simpler
rules and leaves room for numerous improvements on top of our
minimal implementation.

\paragraph{Deduction Rules.}

>>>>

\section{Syntax and Semantics}
\label{sec:sem}

@@

\paragraph{Syntax.}

@@

\paragraph{Semantics.}

@@

We generalize classic Hoare logic to express not only classical
boolean-valued assertions but also assertions that talk about the
future resource usage.  Instead of the usual assertions $P : \State
\to \mathit{bool}$ of Hoare logic we use assertions
$$
P : \State \to \Q \cup \{ \infty \} \; .
$$
This can be understood as a refinement of boolean assertions where
$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Q$.
We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
$\lambda \state . \, \infty$.  We sometimes call assertions
\emph{potential functions}.  To use Coq's support for propositional
reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
in the implementation.  For a given $\state \in \State$, such an
assertion can be seen as a set $B \subseteq \Q$ of valid bounds.
However, we find the presentation in this article easier to read.

@@

\paragraph{Rules of the Quantitative Logic.}

@@

\paragraph{Logical Variables and Shallow Embedding.}

@@

In this section we present the automatic amortized analysis that we
use in our implementation to derive linear resource bounds for C
programs.

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense) for
potential functions.  This allows us to uniquely represent any linear
potential function $\Phi$ as a \emph{quantitative annotation} $Q =
(q_i)_{i \in I}$, that is, a family of rational numbers where only a
finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

In the potential functions, we treat constants as global variables
that can not be assigned to.  For example, if the program contains the
constant $8128$ then we have a variable $c_{8128}$ and
$\state(c_{8128}) = 8128$.  We assume that every program state
includes the constant $c_0$.

@@

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal logical state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can often be inferred by local reasoning (i.e., in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
%
Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we designed and found
in the literature.

\paragraph{Inference Rules.}
